\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-07 chapters/bttda.tex
checking bttda

Due to its multichannel time series structure, EEG data, like most neural
signal acquisition modalities used for BCIs, naturally exist as multiway data,
capturing information in both spatial and temporal domains.
Common preprocessing transformations, such as time-frequency transformation,
time-binning, or integrating information across multiple subjects or conditions,
can further expand the data into additional analytic domains.
This can result in high-dimensional datasets which are usually flattened into a
set of sample vectors, stripping the original data from its structure.
Yet, the intrinsic multiway structure of neural data~\cite{Erol2022} is
well-suited for representation as \emph{tensors}, or multiway arrays, where
each domain is represented as a tensor \emph{mode}.
Tensors provide a structured data representation for this highly dimensional
multiway data
This in turn paves the way to the development of tensor methods which can
counteract some of the drawbacks of the dimensionality problem.
Tensor methods are machine learning techniques that  consider each tensor mode
separately, reducing a given problem into partial, per-mode problems.

These tensor methods decompose a tensor into a lower dimensional structure of a
core tensor and factor tensors.
The most common aproaches adhere to either the Tucker structure or the PARAFAC
structure.
A Tucker decomposition reduces an input tensor of size $(D_1,D_2,\ldots,D_K)$ to
a dense  tensor of size $(r_1,r_2,\ldots,r_K)$ with $r_k<=D_k$ using a
set of per mode factor matrices.
Effective unsupervised tensor decomposition and approximation in the Tucker format can be achieved
using the Higher Order Singular Value
Decomposition (HOSVD)~\cite{DeLathauwer2000,SoleCasals2018}.
Alternatively, the PARAFAC structure can be used.
Here, the tensor is decomposed into a sum of rank-1 tensors, each the product
of a scalar and a vector per mode.
This is equivalent to a Tucker structured decomposition with all core elements
off the hyperdiagonal set to 0.
One way of obtaining an unsupervised PARAFAC decomposition is through the Canonical Polyadic
Decomposition~\cite{Hitchcock1927,Nazarpour2006}.

While commonly used, these Tucker or PARAFAC structure might still not be able to
efficiently represent relevant neural information in a compressed format.
The block-term tensor structure is a generalization of the Tucker and
PARAFAC structures.
It represents the tensor as a sum of Tucker structured terms.
If the number of terms is equal to 1, it is equivalent to the Tucker structure;
if the rank of each term is equal to 1, it is equivalent to the PARAFAC
structure.
The generalized structure can be calculated in an unsupervised way using
the Block-Term Decomposition~\cite{DeLathauwer2008,DeLathauwer2008a,DeLathauwer2008b,Rontogiannis2021}.
Performance of methods leveraging either the Tucker and PARAFAC structure are
heavily dependent on the a priori choice of hyperparameters describing the
respectively the multilinear rank or the number of rank-1 terms.
The block-term structure is more flexible than either the Tucker or the PARAFAC
structures, since it is not constrained to problems that cannot be expressed by
one of these structures and the chosen hyperparameters.
However, this increased flexibility also increased the number of
hyperparameters to both the number of terms and the multilinear rank of each term.

@@

Variants of HODA have been applied to BCI problems such as
ERP~\cite{Onishi2012,Higashi2016} and motor imagery (MI)~\cite{Liu2015,Cai2021}
decoding with positive results~\cite{Lotte2018}.
Recent work proposes adaptations such as suited objective
functions and regularization~\cite{Jamshidi2017,Jorajuria2022,Aghili2023}.
Discriminant tensor features have also be extracted
in the PARAFAC structure through manifold optimization~\cite{Froelich2018}.
However, it is not immediately obvious if either the Tucker or PARAFAC
structure are most suited to represent the neural data of interest for the \ac{bci}
paradigm and for decoding.

Recent research has shown that supervised decoders adopting a more flexible structure
can improve BCI performance.
Promising results have been achieved for regression tasks using
Higher Order Parial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
Regression (BTTR)~\cite{Faes2022,Faes2022b}.
BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
but this leaving room for improvement.
Instead of optimizing features directly for class separability, rather a dummy
independent variable was regressed towards, and the method
cannot be extended to a multi-class setting.
Furthermore, structures employed in these regression are still more constrained
than what could be achieved with a full block-term tensor structured decomposition
optimized for discriminability, since they rely on a low rank common subspace
between the input and classification labels.
\textcite{Huang2020} propose a supervized approach for finding multiple discriminant
multilinear spectral filter terms and apply it to motor imagery \ac{bci}, but their
decomposition is also restricted in flexibility.

\todo{It's always trichiin in the related work section of a ML/Mathematical
paper, but unless you r reviewers are absolute experts of this specific field,
you should at least try to better describe the previous methods.}

@@

A block-term decomposition that is directly optimized for discriminability and with a
good choice of ranks and number of terms might better represent the behaviour
of generators of neural activity through increased flexibility, or might
achieve better regularization through increased sparsity.
An alternative view on the same approach goes as follows:
If HODA with a well chosen multilinear rank extracts some discriminant features
from the input tensor, it is likely that it does not yet retrieve all useful
information due to the restriction following from its Tucker structure.
Could we therefore not keep using HODA to extract discriminant Tucker structured
as long as decoding performance increases?

We propose to implement this idea as a new supervised feature
extraction tensor method that is a generalization of the aforementioned HODA
algorithm and is more suited for the extraction of discriminant
features while adhering to a flexible and efficient block-term tensor
structure.
This work features the following contributions:
\begin{enumerate*}[label={\arabic*)}]
	\item  We develop a forward model for HODA to reconstruct a
	      given input tensor from the extracted features.
	\item This allows us to introduce a state-of-the-art BCI feature extraction
        method based on the block-term tensor structure, named Block-Term Tensor Discriminant Analysis
	      (BTTDA).
  \item We evaluate a BCI decoder based on BTTDA and its special
	      PARAFAC-structured case on decoding tasks for both ERP and MI
	      paradigm BCI datasets.
\end{enumerate*}
\section{Methods}

\subsection{Notation}
Tensors are indicated by bold underlined letters $\ten{X}$, matrices by bold
letters $\mat{U}$, fixed scalars by uppercase letters $K$ and variable
scalars as lower case letters $k$.
The $n^\text{th}$ sample of a tensor dataset with $N$ samples is written as
$\ten{X}(n)$.
A tensor $\ten{X}\in \mathbb{R}^{D_1\times D_2 \times \cdots \times D_K}$ can be
unfolded in mode $k$ to a matrix
$\mat{X}_k\in\mathbb{R}^{(D_k\times\prod_{j\neq k}^K D_j)}$, by concatenating
all mode $j\neq$ fibers.
The tensor-matrix product of tensor $\ten{X}$ with matrix $\mat{U}$ along a
given mode $k$ is written as $\ten{X}\mpr{\mat{U}}{k}$. For ease of notation, let
$\ten{X}\mmpr{\mat{U}} =
	\ten{X}\mpr{\mat{U}}{1}\mpr{\mat{U}}{2}\cdots\mpr{\mat{U}}{K}$.
When skipping one of the modes $k$, this is
written as $\ten{X}\mmprs{\mat{U}}{k} =
	\ten{X}\mpr{\mat{U}}{1}\mpr{\mat{U}}{2}\cdots\mpr{\mat{U}}{k-1}\mpr{\mat{U}}{k+1}\ldots\mpr{\mat{U}}{K}$.
%The Kronecker product is noted as $\otimes$, covariance matrices are indicated  with $\mat{\Sigma}$.

\subsection{Higher Order Discriminant Analysis}
Higher Order Discriminant Analysis (HODA)~\cite{Phan2010} is a
supervised tensor-based feature extraction technique.
For a set of $N$ tensors of order $K$
$\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
		D_K}\right\}_n^N$, HODA finds projection matrices $\mat{U_k}$ for each mode $k$
that project a given $\ten{X}$ to a latent tensor
$\ten{G}\in\mathbb{R}^{r_1\times r_2\times\cdots\times r_K}$, usually with lower
dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
tensor-matrix mode products
\begin{equation}
	\ten{G}  = \ten{X}\mmpr{\mat{U}}
	\label{eq:HODA-backward}
\end{equation}
visualized in Figure~\ref{fig:HODA-backward}.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_backward.tikz.tex}
  \caption[The \acs{hoda} backward projection.]{%
    A visualization of the multilinear projection obtained by Higher Order
    Discriminant Analysis (HODA) applied to a third-order tensor
    sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
		HODA finds projection matrices $\mat{U}_k$ such that maximal
		discriminability between classes is achieved in the projected latent tensors
		$\ten{G}$ with reduced dimensionality $(r_1,r_2,r_3)$.}
	\label{fig:HODA-backward}
\end{figure}
Analogous to the HOSVD, HODA is a dimensionality
reduction decomposition that results in a dense latent tensor $\ten{G}$, and
imposes an orthogonality constraint on $\mat{U}_k$ to ensure uniqueness.
However, while for the HOSVD decomposition the projection matrices
are chosen to minimize the reconstruction error, the projection matrices
$\mat{U}_k$ of HODA are optimized for maximal discriminability of tensors
$\ten{G}(n)$ belonging to classes with labels $c_n$.
This is a desirable property in a classification setting where samples are
high-dimensional tensors.

HODA optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent
tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which is solved through the HODA algorithm.
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per mode Singular Value Decomposition (SVD),
or as the partial HOSVD of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
%and solving the generalized eigenvalue problem
%\begin{equation}
%  = \mat{\lambda}\mat{S}_{-k,\text{w}}\mat{U}_k
%\end{equation}
The iterative process halts after a fixed amount or iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.
The full HODA procedure is summarized in Algorithm~\ref{alg:HODA}.
\begin{algorithm}
  \caption[The \acs{hoda} backward solution.]{The \ac{hoda} backward solution.}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}

@@

To avoid overfitting and improve performance in low sample size settings, the
HODA problem can be regularized by shrinking the partial
within-class scatter matrices~\cite{Phan2010} with a shrinkage factor
$\alpha_k$ at each step such that the eigenvalue problem becomes
\begin{equation}
	\mat{S}_b^{(-k)} -
	\varphi\left[\left(1-\alpha_k\right)\mat{S}_{-k,\text{w}}+\alpha_k\mat{I}\right] =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
As in Linear Discriminant Analysis, the shrinkage parameter $\alpha_k$ can
also be estimated in a data driven way in HODA~\cite{Jorajuria2022},
e.g., using the Ledoit-Wolf procedure~\cite{Ledoit2003} at every iteration.

\subsection{A forward model for HODA}
As a prerequisite to the proposed BTTDA model, we must find a
way to reconstruct the original data tensor $\ten{X}$ as accurately as possible
from $\ten{G}$ after dimensionality reduction.
This is usually referred to as finding a corresponding \emph{forward model}.
While a \emph{backward model} extracts latent sources or properties from the observed
data based on some task related criterion or on prior domain knowledge,
a forward model is a generative model that expresses the observed data in
function of these latent properties or sources that are given.
Forward models are useful for a.o.\ interpretability and data compression.

The HODA projection in Equation~\ref{eq:HODA-backward} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model, visualized in Figure~\ref{fig:HODA-forward} is given by
\begin{equation}
  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
  \hat{\ten{X}} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
reconstructed tensor $\hat{\ten{X}}$ and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA.
    By calculating activation	patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
    the error term $\ten{\mathbfcal{E}}$.}
	\label{fig:HODA-forward}
\end{figure}

>>>>

Due to its multichannel time series structure, EEG data, like most neural
signal acquisition modalities used for BCIs, naturally exist as multiway data,
capturing information in both spatial and temporal domains.
Common preprocessing transformations, such as time-frequency transformation,
time-binning, or integrating information across multiple subjects or conditions,
can further expand the data into additional analytic domains.
This can result in high-dimensional datasets which are usually flattened into a
set of sample vectors, stripping the original data from its structure.
Yet, the intrinsic multiway structure of neural data~\cite{Erol2022} is
well-suited for representation as \emph{tensors}, or multiway arrays, where
each domain is represented as a tensor \emph{mode}.
Tensors provide a structured data representation for this highly dimensional
multiway data.
This in turn paves the way to the development of tensor methods which can
counteract some of the drawbacks of the dimensionality problem.
Tensor methods are machine learning techniques that consider each tensor mode
separately, reducing a given problem into partial, per-mode problems.

These tensor methods decompose a tensor into a lower dimensional structure of a
core tensor and factor tensors.
The most common approaches adhere to either the Tucker structure or the PARAFAC
structure.
A Tucker decomposition reduces an input tensor of size $(D_1,D_2,\ldots,D_K)$ to
a dense tensor of size $(r_1,r_2,\ldots,r_K)$ with $r_k \leq D_k$ using a
set of per-mode factor matrices.
Effective unsupervised tensor decomposition and approximation in the Tucker format can be achieved
using the Higher Order Singular Value
Decomposition (HOSVD)~\cite{DeLathauwer2000,SoleCasals2018}.
Alternatively, the PARAFAC structure can be used.
Here, the tensor is decomposed into a sum of rank-1 tensors, each the product
of a scalar and a vector per mode.
This is equivalent to a Tucker structured decomposition with all core elements
off the hyperdiagonal set to 0.
One way of obtaining an unsupervised PARAFAC decomposition is through the Canonical Polyadic
Decomposition~\cite{Hitchcock1927,Nazarpour2006}.

While commonly used, these Tucker or PARAFAC structures might still not be able to
efficiently represent relevant neural information in a compressed format.
The block-term tensor structure is a generalization of the Tucker and
PARAFAC structures.
It represents the tensor as a sum of Tucker structured terms.
If the number of terms is equal to 1, it is equivalent to the Tucker structure; if the rank of each term is equal to 1, it is equivalent to the PARAFAC
structure.
The generalized structure can be calculated in an unsupervised way using
the Block-Term Decomposition~\cite{DeLathauwer2008,DeLathauwer2008a,DeLathauwer2008b,Rontogiannis2021}.
Performance of methods leveraging either the Tucker and PARAFAC structures are
heavily dependent on the a priori choice of hyperparameters describing
the multilinear rank or the number of rank-1 terms.
The block-term structure is more flexible than either the Tucker or the PARAFAC
structures, since it is not constrained to problems that cannot be expressed by
one of these structures and the chosen hyperparameters.
However, this increased flexibility also increases the number of
hyperparameters to both the number of terms and the multilinear rank of each term.

@@

Variants of HODA have been applied to BCI problems such as
ERP~\cite{Onishi2012,Higashi2016} and motor imagery (MI)~\cite{Liu2015,Cai2021}
decoding with positive results~\cite{Lotte2018}.
Recent work proposes adaptations such as suited objective
functions and regularization~\cite{Jamshidi2017,Jorajuria2022,Aghili2023}.
Discriminant tensor features have also been extracted
in the PARAFAC structure through manifold optimization~\cite{Froelich2018}.
However, it is not immediately obvious if either the Tucker or PARAFAC
structure are most suited to represent the neural data of interest for the BCI
paradigm and for decoding.

Recent research has shown that supervised decoders adopting a more flexible structure
can improve BCI performance.
Promising results have been achieved for regression tasks using
Higher Order Partial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
Regression (BTTR)~\cite{Faes2022,Faes2022b}.
BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
but this leaves room for improvement.
Instead of optimizing features directly for class separability, a dummy
independent variable was regressed towards, and the method
cannot be extended to a multi-class setting.
Furthermore, structures employed in these regression approaches are still more constrained
than what could be achieved with a full block-term tensor structured decomposition
optimized for discriminability, since they rely on a low-rank common subspace
between the input and classification labels.
\textcite{Huang2020} propose a supervised approach for finding multiple discriminant
multilinear spectral filter terms and apply it to motor imagery BCI, but their
decomposition is also restricted in flexibility.

\todo{It's always tricky in the related work section of an ML/Mathematical
paper, but unless your reviewers are absolute experts of this specific field,
you should at least try to better describe the previous methods.}

@@

A block-term decomposition that is directly optimized for discriminability and with a
good choice of ranks and number of terms might better represent the behavior
of generators of neural activity through increased flexibility, or might
achieve better regularization through increased sparsity.
An alternative view on the same approach goes as follows:
If HODA with a well-chosen multilinear rank extracts some discriminant features
from the input tensor, it is likely that it does not yet retrieve all useful
information due to the restriction following from its Tucker structure.
Could we therefore not keep using HODA to extract discriminant Tucker structured
terms as long as decoding performance increases?

We propose to implement this idea as a new supervised feature
extraction tensor method that is a generalization of the aforementioned HODA
algorithm and is more suited for the extraction of discriminant
features while adhering to a flexible and efficient block-term tensor
structure.
This work features the following contributions:
\begin{enumerate*}[label={\arabic*)}]
	\item  We develop a forward model for HODA to reconstruct a
	      given input tensor from the extracted features.
	\item This allows us to introduce a state-of-the-art BCI feature extraction
        method based on the block-term tensor structure, named Block-Term Tensor Discriminant Analysis
	      (BTTDA).
  \item We evaluate a BCI decoder based on BTTDA and its special
	      PARAFAC-structured case on decoding tasks for both ERP and MI
	      paradigm BCI datasets.
\end{enumerate*}

\section{Methods}

\subsection{Notation}
Tensors are indicated by bold underlined letters $\ten{X}$, matrices by bold
letters $\mat{U}$, fixed scalars by uppercase letters $K$, and variable
scalars as lowercase letters $k$.
The $n^\text{th}$ sample of a tensor dataset with $N$ samples is written as
$\ten{X}(n)$.
A tensor $\ten{X}\in \mathbb{R}^{D_1\times D_2 \times \cdots \times D_K}$ can be
unfolded in mode $k$ to a matrix
$\mat{X}_k\in\mathbb{R}^{(D_k\times\prod_{j\neq k}^K D_j)}$, by concatenating
all mode $j\neq k$ fibers.
The tensor-matrix product of tensor $\ten{X}$ with matrix $\mat{U}$ along a
given mode $k$ is written as $\ten{X}\mpr{\mat{U}}{k}$. For ease of notation, let
$\ten{X}\mmpr{\mat{U}} =
	\ten{X}\mpr{\mat{U}}{1}\mpr{\mat{U}}{2}\cdots\mpr{\mat{U}}{K}$.
When skipping one of the modes $k$, this is
written as $\ten{X}\mmprs{\mat{U}}{k} =
	\ten{X}\mpr{\mat{U}}{1}\mpr{\mat{U}}{2}\cdots\mpr{\mat{U}}{k-1}\mpr{\mat{U}}{k+1}\ldots\mpr{\mat{U}}{K}$.

\subsection{Higher Order Discriminant Analysis}
Higher Order Discriminant Analysis (HODA)~\cite{Phan2010} is a
supervised tensor-based feature extraction technique.
For a set of $N$ tensors of order $K$
$\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
		D_K}\right\}_n^N$, HODA finds projection matrices $\mat{U_k}$ for each mode $k$
that project a given $\ten{X}$ to a latent tensor
$\ten{G}\in\mathbb{R}^{r_1\times r_2\times\cdots\times r_K}$, usually with lower
dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
tensor-matrix mode products:
\begin{equation}
	\ten{G}  = \ten{X}\mmpr{\mat{U}}
	\label{eq:HODA-backward}
\end{equation}
as visualized in \autoref{fig:HODA-backward}.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_backward.tikz.tex}
  \caption[A \acs{hoda} backward projection.]{%
    A visualization of the multilinear projection obtained by Higher Order
    Discriminant Analysis (HODA) applied to a third-order tensor
    sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
		HODA finds projection matrices $\mat{U}_k$ such that maximal
		discriminability between classes is achieved in the projected latent tensors
		$\ten{G}$ with reduced dimensionality $(r_1,r_2,r_3)$.}
	\label{fig:HODA-backward}
\end{figure}

Analogous to the HOSVD, HODA is a dimensionality
reduction decomposition that results in a dense latent tensor $\ten{G}$, and
imposes an orthogonality constraint on $\mat{U}_k$ to ensure uniqueness.
However, while for the HOSVD decomposition the projection matrices
are chosen to minimize the reconstruction error, the projection matrices
$\mat{U}_k$ of HODA are optimized for maximal discriminability of tensors
$\ten{G}(n)$ belonging to classes with labels $c_n$.
This is a desirable property in a classification setting where samples are
high-dimensional tensors.

HODA optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$:
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices:
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which is solved through the HODA algorithm.
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per-mode Singular Value Decomposition (SVD),
or as the partial HOSVD of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor:
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem:
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices:
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of:
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
The iterative process halts after a fixed number of iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.
The full HODA procedure is summarized in \autoref{alg:HODA}.
\begin{algorithm}
  \caption[A \acs{hoda} backward solution.]{The \acs{hoda} backward solution.}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}

@@

To avoid overfitting and improve performance in low sample size settings, the
HODA problem can be regularized by shrinking the partial
within-class scatter matrices~\cite{Phan2010} with a shrinkage factor
$\alpha_k$ at each step such that the eigenvalue problem becomes:
\begin{equation}
	\mat{S}_b^{(-k)} -
	\varphi\left[\left(1-\alpha_k\right)\mat{S}_{-k,\text{w}}+\alpha_k\mat{I}\right] =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
As in Linear Discriminant Analysis, the shrinkage parameter $\alpha_k$ can
also be estimated in a data-driven way in HODA~\cite{Jorajuria2022},
e.g., using the Ledoit-Wolf procedure~\cite{Ledoit2003} at every iteration.

\subsection{A forward model for HODA}
As a prerequisite to the proposed BTTDA model, we must find a
way to reconstruct the original data tensor $\ten{X}$ as accurately as possible
from $\ten{G}$ after dimensionality reduction.
This is usually referred to as finding a corresponding \emph{forward model}.
While a \emph{backward model} extracts latent sources or properties from the observed
data based on some task-related criterion or on prior domain knowledge,
a forward model is a generative model that expresses the observed data in
terms of these latent properties or sources that are given.
Forward models are useful for, among other things, interpretability and data compression.

The HODA projection in \autoref{eq:HODA-backward} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model, visualized in \autoref{fig:HODA-forward}, is given by:
\begin{equation}
  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
  \hat{\ten{X}} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
reconstructed tensor $\hat{\ten{X}}$, and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[A forward projection for \ac{hoda}.]{The forward projection for HODA.
    By calculating activation patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
    the error term $\ten{\mathbfcal{E}}$.}
	\label{fig:HODA-forward}
\end{figure}

A forward model should ensure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should be maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors:
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n)
  -\hat{\ten{X}}(n)\right)^2
  = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmprs{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares (ALS) algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed number of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in \autoref{alg:HODA-fw}.
\begin{algorithm}
  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

The forward model could alternatively be specified in the classical manner by linear
regression or with the solution of \textcite{Haufe2014}, but this would require
vectorizing the tensor representation.
While the reconstruction error might be reduced, we would lose the computational
efficiency and regularizing constraints of the tensor form.
More efficiently, the forward model could be obtained through other tensor
regression methods, like Higher Order Partial
Least Squares.
This method still requires fitting more parameters than our
proposed model, on top of additional hyperparameters defining the
dimensionality of the common subspace.
An attractive property of our multilinear forward model for HODA is that it
estimates exactly as many parameters as the backward model, with the intuition that
it should not disproportionally contribute to overfitting.
